---
title: "Data Preprocessing"
output: 
    html_document:
        toc: TRUE   # Table content at front of the page
        toc_float: TRUE   # Floating table content on the left
---

```{r}
#### Workspace setup ####
library(tidyverse)
library(janitor)
library(dplyr)
library(readxl)
library(readr)
library(hms)

#### Read data ####
bus_delay_2024 <- read.csv("data/raw/bus-data.csv")
streetcar_delay_2024 <- read.csv("data/raw/streetcar-data.csv")
subway_delay_2024 <- read.csv("data/raw/subway-data.csv")
delay_codes <- read.csv("data/raw/subway-delay-codes.csv")

#### Clean data ####
###Delay Data ###
ttc_bus_delay <-
  bus_delay_2024 |>
  janitor::clean_names() |>  
  mutate(date = dmy(date))|>
  select(date, route, time, day, location, incident, min_delay, min_gap, direction)

ttc_streetcar_delay <-
  streetcar_delay_2024 |>
  janitor::clean_names() |>
  mutate(date = dmy(date))|>
  select(date, line, time, day, location, incident, min_delay, min_gap, bound)

ttc_subway_delay <-
  subway_delay_2024 |>
  janitor::clean_names() |>
  mutate(date = ymd(date))|>
  select(date, time, day, station, code, min_delay, min_gap, bound, line)

```

```{r}
### Filter Dataset for weekday and weekend ###

## Weekday vs Weekend Delay
ttc_bus_delay <- ttc_bus_delay |>
  mutate(day_type = ifelse(day %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"), 
                           "Weekday", "Weekend"))
ttc_streetcar_delay <- ttc_streetcar_delay |>
  mutate(day_type = ifelse(day %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"), 
                           "Weekday", "Weekend"))
ttc_subway_delay <- ttc_subway_delay |>
  mutate(day_type = ifelse(day %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"), 
                           "Weekday", "Weekend"))
```

```{r}
### Filter Dataset for rush-Hour Delay
# Bus
ttc_bus_delay <- ttc_bus_delay |>
  mutate(
    time = as_hms(paste0(time, ":00")),  # Convert 'time' to HH:MM:SS format
    rush_type = case_when(
      time >= as_hms("06:00:00") & time <= as_hms("09:00:00") ~ "morning",  # Morning rush (06:00–09:00)
      time >= as_hms("15:00:00") & time <= as_hms("19:00:00") ~ "evening",  # Evening rush (15:00–19:00)
      TRUE ~ "none"  # Not in rush hours
    )
  )

# Streetcar
ttc_streetcar_delay <- ttc_streetcar_delay |>
  mutate(
    time = as_hms(paste0(time, ":00")),  # Convert 'time' to HH:MM:SS format
    rush_type = case_when(
      time >= as_hms("06:00:00") & time <= as_hms("09:00:00") ~ "morning",  # Morning rush (06:00–09:00)
      time >= as_hms("15:00:00") & time <= as_hms("19:00:00") ~ "evening",  # Evening rush (15:00–19:00)
      TRUE ~ "none"  # Not in rush hours
    )
  )

# Subway
ttc_subway_delay <- ttc_subway_delay |>
  mutate(
    time = as_hms(paste0(time, ":00")),  # Convert 'time' to HH:MM:SS format
    rush_type = case_when(
      time >= as_hms("06:00:00") & time <= as_hms("09:00:00") ~ "morning",  # Morning rush (06:00–09:00)
      time >= as_hms("15:00:00") & time <= as_hms("19:00:00") ~ "evening",  # Evening rush (15:00–19:00)
      TRUE ~ "none"  # Not in rush hours
    )
  )
```

```{r}
### Filter Dataset for seasons ###
#### Bus
ttc_bus_delay <- ttc_bus_delay |>
  janitor::clean_names() |>
  mutate(
    date = as.Date(date),  # Ensure the 'date' is in Date format
    season = case_when(
      month(date) %in% c(3, 4, 5) ~ "Spring",    # March, April, May
      month(date) %in% c(6, 7, 8) ~ "Summer",    # June, July, August
      month(date) %in% c(9, 10, 11) ~ "Fall",    # September, October, November
      month(date) %in% c(12, 1, 2) ~ "Winter",   # December, January, February
      TRUE ~ "Unknown"  # In case there are any issues
    )
  )

#### Streetcar
ttc_streetcar_delay <- ttc_streetcar_delay |>
  janitor::clean_names() |>
  mutate(
    date = as.Date(date),
    season = case_when(
      month(date) %in% c(3, 4, 5) ~ "Spring",
      month(date) %in% c(6, 7, 8) ~ "Summer",
      month(date) %in% c(9, 10, 11) ~ "Fall",
      month(date) %in% c(12, 1, 2) ~ "Winter",
      TRUE ~ "Unknown"
    )
  )

#### Subway
ttc_subway_delay <- ttc_subway_delay |>
  janitor::clean_names() |>
  mutate(
    date = as.Date(date),
    season = case_when(
      month(date) %in% c(3, 4, 5) ~ "Spring",
      month(date) %in% c(6, 7, 8) ~ "Summer",
      month(date) %in% c(9, 10, 11) ~ "Fall",
      month(date) %in% c(12, 1, 2) ~ "Winter",
      TRUE ~ "Unknown"
    )
  )

```

```{r}
#### Filter Dataset for Holidays
#https://www.toronto.ca/home/contact-us/statutory-holidays/

# Holidays for 2024
holidays_2024 <- c(
  "2024-01-01",  # New Year’s Day
  "2024-02-19",  # Family Day
  "2024-03-29",  # Good Friday
  "2024-04-01",  # Easter Monday
  "2024-05-20",  # Victoria Day
  "2024-07-01",  # Canada Day
  "2024-08-05",  # Simcoe (Civic) Day
  "2024-09-02",  # Labour Day
  "2024-10-14",  # Thanksgiving Day
  "2024-11-11",  # Remembrance Day
  "2024-12-25",  # Christmas Day
  "2024-12-26"   # Boxing Day
)

# Convert holiday dates to Date class
holidays_2024 <- as.Date(holidays_2024)

# Add 'is_holiday' column to each dataset
ttc_bus_delay <- ttc_bus_delay |>
  janitor::clean_names() |>
  mutate(
    date = as.Date(date),  # Ensure date is in Date format
    is_holiday = ifelse(date %in% holidays_2024, 1, 0)  # 1 for holiday, 0 for non-holiday
  )

ttc_streetcar_delay <- ttc_streetcar_delay |>
  janitor::clean_names() |>
  mutate(
    date = as.Date(date),
    is_holiday = ifelse(date %in% holidays_2024, 1, 0)
  )

ttc_subway_delay <- ttc_subway_delay |>
  janitor::clean_names() |>
  mutate(
    date = as.Date(date),
    is_holiday = ifelse(date %in% holidays_2024, 1, 0)
  )

```

```{r}
#### Data Cleaning
##Missing Values
ttc_bus_delay <- ttc_bus_delay %>%
  filter(!is.na(date) & !is.na(time) & !is.na(location))

ttc_streetcar_delay <- ttc_streetcar_delay %>%
  filter(!is.na(date) & !is.na(time) & !is.na(location))

ttc_subway_delay <- ttc_subway_delay %>%
  filter(!is.na(date) & !is.na(time) & !is.na(station))

```

```{r}
### Adding weather
library(tidyverse)


#### weather data
toronto_weather_all_data <- read_csv("data/raw/en_climate_daily_ON_6158355_2024_P1D.csv")


toronto_weather <- toronto_weather_all_data |>
  select(`Date/Time`, Year, Month, Day, `Max Temp (°C)`, `Min Temp (°C)`, `Total Precip (mm)`, `Snow on Grnd (cm)`) |>
  mutate("date" = `Date/Time`) |>
  select(-`Date/Time`)


#### merging ttc data with weather data
ttc_bus_delay <- merge(ttc_bus_delay, toronto_weather, by = "date")


ttc_streetcar_delay <- merge(ttc_streetcar_delay, toronto_weather, by = "date")


ttc_subway_delay <- merge(ttc_subway_delay, toronto_weather, by = "date")
```

# Subway Preprocessing
For simplicity, only keep the delays that are at one station (not in between or multiple stations)
```{r}
subway_subset <- ttc_subway_delay %>%
  filter(str_detect(station, "STATION")) 
```

```{r}
dim(subway_subset) # nice! most of them are kept
```

Clean the text for the stations so I can join. 
```{r}
subway_subset$station <- str_remove(subway_subset$station, " STATION$") # remove STATION
subway_subset$station <- str_remove(subway_subset$station, " BD$") # remove BD
subway_subset$station <- str_remove(subway_subset$station, " YUS$") # remove YUS
subway_subset$station <- str_to_title(subway_subset$station) # change to title case
```

```{r}
sort(table(subway_subset$station), decreasing = TRUE) # weird names that will get messed up
```
There's some stations here that will definitely get lost in the join with the ESRI data -- some stations have been renamed/merged, some formatting/abbreviation differences, and some are misspellings. For example, Spadina YUS Station, Bloor Station, St Geoge (?!)

Manually rename these subway stations to match the ESRI data
```{r}
subway_subset <- subway_subset %>% 
  mutate(station = recode(station,
                          "Vaughan Mc" = "Vaughan Met. Centre",
                          "North York Ctr" = "North York Centre",
                          "St Geoge" = "St George",
                          "St. George" = "St George",
                          "Kenndy" = "Kennedy",
                          "Queens Park" = "Queen's Park",
                          "Old Mills" = "Old Mill",
                          "St. Andrew" = "St Andrew",
                          "Spadina Yu" = "Spadina",
                          "Sheppard Yonge" = "Sheppard",
                          "Wislon" = "Wilson",
                          "Sheppard-Yonge" = "Sheppard",
                          "Main Street" = "Main"))
```

In ArcGIS Pro, I imported ESRI Canada Education's [Toronto Subway Stations point data](https://www.arcgis.com/home/item.html?id=05200e06ff524319bde9f16e5955496b). 
Then do the following geoprocessing to obtain `ttc-station-point.csv`: Geodatabase to Shape, Add XY Coordinates, Export Table.
```{r}
# grab nice TTC Station coordinates from ESRI ArcGIS
ttc_coords <- read_csv("data/raw/ttc-station-point.csv")
```

Even though Bloor and Yonge are in the same location (and same point on the ESRI data), let's add them separately since the delays on the different lines are somewhat independent
```{r}
ttc_coords <- ttc_coords %>%
  # Filter the "Bloor-Yonge" row
  bind_rows(
    ttc_coords %>%
      filter(station == "Bloor-Yonge") %>%
      mutate(station = "Bloor",
             line = "Bloor-Danforth",
             line_1 = 2,
             POINT_Y = 43.6700, # slightly offset so they can be distinguished on the plot
             POINT_X = -79.3852
             ),
    ttc_coords %>%
      filter(station == "Bloor-Yonge") %>%
      mutate(station = "Yonge")
  ) %>%
  # remove the original "Bloor-Yonge" row
  filter(station != "Bloor-Yonge")

ttc_coords
```

Merge the XY coordinates to the delay data
```{r}
delay_coords <- left_join(subway_subset, ttc_coords, by = "station")
delay_coords
```

See what other stations are mismatched and continue to iterate and rerun
```{r}
mismatched <- delay_coords %>% filter(is.na(POINT_X))
sort(table(mismatched$station), decreasing = TRUE)
```
OK. Great, it seems like it is all the multiple stations / between stations now. So, let's remove them and now we can continue with the join!! 
```{r}
matched <- delay_coords %>% filter(!is.na(POINT_X))
```

Merge with the delay codes
```{r}
delay_codes <- delay_codes %>% janitor::clean_names()
ttc_subway_delay <- left_join(matched, delay_codes, by = join_by(code == sub_rmenu_code))
ttc_subway_delay <- ttc_subway_delay %>% select(!number)
ttc_subway_delay
```

# Dealing with Missing Values
```{r}
ttc_bus_delay$direction[ttc_bus_delay$direction == ""] <- NA
sum(is.na(ttc_bus_delay$direction))
ttc_streetcar_delay$bound[ttc_streetcar_delay$bound == ""] <- NA
sum(is.na(ttc_streetcar_delay$bound))
ttc_subway_delay$bound[ttc_subway_delay$bound == ""] <- NA
sum(is.na(ttc_subway_delay$bound))

# Check missing values
na_counts <- colSums(is.na(ttc_bus_delay))
na_counts[na_counts > 0]

# Replace NA values in 'Snow on Grnd (cm)' with 0
ttc_bus_delay$`Snow on Grnd (cm)`[is.na(ttc_bus_delay$`Snow on Grnd (cm)`)] <- 0
ttc_streetcar_delay$`Snow on Grnd (cm)`[is.na(ttc_streetcar_delay$`Snow on Grnd (cm)`)] <- 0
ttc_subway_delay$`Snow on Grnd (cm)`[is.na(ttc_subway_delay$`Snow on Grnd (cm)`)] <- 0

ttc_bus_delay <- ttc_bus_delay[order(ttc_bus_delay$date), ]
ttc_streetcar_delay <- ttc_streetcar_delay[order(ttc_streetcar_delay$date), ]
ttc_subway_delay <- ttc_subway_delay[order(ttc_subway_delay$date), ]

# Replace NA in temp and total precip with average of nearest dates
fill_na_with_avg <- function(x) {
  for (i in which(is.na(x))) {
    previous_value <- x[max(which(!is.na(x[1:i-1])))]
    next_value <- x[min(which(!is.na(x[(i+1):length(x)])) + i)]
    x[i] <- mean(c(previous_value, next_value), na.rm = TRUE)
  }
  return(x)
}

for (col in c("Max Temp (°C)", "Min Temp (°C)", "Total Precip (mm)")) {
  ttc_bus_delay[[col]] <- fill_na_with_avg(ttc_bus_delay[[col]])
  ttc_streetcar_delay[[col]] <- fill_na_with_avg(ttc_streetcar_delay[[col]])
  ttc_subway_delay[[col]] <- fill_na_with_avg(ttc_subway_delay[[col]])
}

# Clean Subway
ttc_subway_delay <- ttc_subway_delay |> select(-c("Year", "Month", "Day", "OID_")) |>
  rename("line_code" = "line.x", "line_name" = "line.y", "line_number" = "line_1")
ttc_subway_delay$line_number[ttc_subway_delay$line_number == 0] <- 1
```
```{r}
ttc_bus_delay <-
  ttc_bus_delay |> 
  janitor::clean_names()

ttc_streetcar_delay <-
  ttc_streetcar_delay |> 
  
  janitor::clean_names()

ttc_subway_delay <-
  ttc_subway_delay |> 
  janitor::clean_names()
```


```{r}
#### Save data ####
write_csv(ttc_bus_delay, "data/cleaned/cleaned_bus.csv")
write_csv(ttc_streetcar_delay, "data/cleaned/cleaned_streetcar.csv")
write_csv(ttc_subway_delay, "data/cleaned/cleaned_subway.csv")
```
