---
title: "Visualization"
output: 
    html_document:
        toc: TRUE   # Table content at front of the page
        toc_float: TRUE   # Floating table content on the left
---

```{r setup, message=FALSE, warning=FALSE, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, eval = T, include  = T, echo=F) 
library(knitr)
library(data.table)
library(tidyverse)
library(kableExtra)
library(dplyr)
library(splines)
library(mgcv)
library(ggstats)
library(cowplot)
library(plotly)
library(hms)
library(sf)
library(janitor)
library(leaflet)
library(scales)
library(gganimate)
library(gifski)
```

```{r checking-file, echo=FALSE}
# Load dataset
bus <- data.table::fread("data/cleaned/cleaned_bus.csv")
streetcar <- data.table::fread("data/cleaned/cleaned_streetcar.csv")
subway <- data.table::fread("data/cleaned/cleaned_subway.csv")
delay_codes <- data.table::fread("data/raw/subway-delay-codes.csv")
delay_codes <- delay_codes %>% janitor::clean_names()
```

## Subway Geographical Hotspots {.tabset}
With the goal of analyzing contributing factors to delays on TTC buses, streetcars and the subway, we decided to conduct exploratory data analysis on the geographic hotspots of subway delays, the top incidence types, the peak times for delays, and the distribution of delay durations for each transport type.

First, we plot the number of delays reported at each subway station in Toronto in 2024. As seen below, it seems like more delays occur at the endpoints of each line, namely at the Kennedy, Kipling, and Finch stations. Moreover, the figure suggests that the highest number of delays occurs at Bloor station, with 1008, although the average delay is only 1.97 minutes. Interacting with the full map, it suggests that disorderly patrons make up a substantial amount of the reasons for delays.

```{r echo=FALSE}
# function to calculate mode
get_mode <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]  # return the most frequent value
}
```

```{r echo=FALSE}
by_station <- subway %>%
  group_by(station) %>%
  summarize(
    avg_delay = mean(min_delay, na.rm = TRUE),
    num_delays = n(),
    lon = first(point_x), # first is ok, because they are all the same for each station
    lat = first(point_y),
    most_frequent_code = get_mode(code),
    freq_count = sum(code == most_frequent_code)
      )

by_station <- by_station %>% 
  left_join(delay_codes, by = join_by(most_frequent_code == sub_rmenu_code))
```

```{r echo=FALSE}
# Read in the shapefile of the TTC Subway lines (from [Toronto Open Data](https://open.toronto.ca/dataset/ttc-subway-shapefiles/)). 
my_sf <- read_sf("data/raw/ttc-subway-shapefile-wgs84/TTC_SUBWAY_LINES_WGS84.shp")
```


```{r}
# Plot the map!!!
pal <- colorNumeric(
  palette = c("salmon", "purple", "blue"),
  # palette = "magma",
  domain = by_station$num_delays
)

leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolylines(data = my_sf, 
               color = ~case_when(
                 RID == 1 ~ "darkgoldenrod",      
                 RID == 2 ~ "darkgreen",    
                 RID == 4 ~ "purple", 
                 # TRUE ~ "gray"  # Default color for other lines if not specified
               ),
               weight = 5, opacity = 0.5) %>% 

  addCircleMarkers(
    data = by_station,
    lng = ~lon, lat = ~lat, 
    color = ~pal(num_delays),
    fillOpacity = 0.8,
    radius = ~rescale(num_delays, to = c(1, 10)),
    popup = ~paste0("<b>", station, "</b><br>Number of Delays: ", num_delays, 
                    "<br> Average Delay: ", round(avg_delay, 2), 
                    " minutes <br> Most Frequent Reason: ", code_description, 
                    " (", freq_count, ")")
  ) %>%
  
  addLegend(pal = pal, 
            values = by_station$num_delays,
            title = "Number of Delays",
            position = 'bottomright') %>% 
  addLegend(
    colors = c("darkgoldenrod", "darkgreen", "purple"),
    labels = c("1", "2", "4"),
    title = "Lines",
    position = 'topright'
  )
```
## Top 10 Incidence Types {.tabset}
Hence, we decide to investigate the most frequent reasons for delays for each transport type, as in the below figures. This is vital to bring insight into how future delays might be mitigated and specific solutions to target. This plot suggests that mechanical reasons are the top reason for delays in buses, whereas operations and security are the top reasons for streetcars. Interestingly, the top reasons for delays in the subway are passenger-related, including disorderly patrons, illness and injury, matching what was previously suggested.

### Bus
```{r echo=FALSE}
# Calculate frequency of each incident
top_10_incidents <- bus |>
  count(incident, sort = TRUE) |>
  top_n(10, n)  # Select the top 10 incidents by frequency

# Plot the histogram for the top 10 incidents
plot_ly(x = reorder(top_10_incidents$incident, -top_10_incidents$n),
        y = top_10_incidents$n, 
        type = "bar") |>
  layout(
    title = "Top 10 Incident Frequency for Bus",
    xaxis = list(title = "Incident"),
    yaxis = list(title = "Frequency")
  )
```

### Streetcar
```{r echo=FALSE}
# Calculate frequency of each incident
top_10_incidents <- streetcar|>
  count(incident, sort = TRUE) |>
  top_n(10, n)  # Select the top 10 incidents by frequency

# Plot the histogram for the top 10 incidents
plot_ly(x = reorder(top_10_incidents$incident, -top_10_incidents$n),
        y = top_10_incidents$n, 
        type = "bar") |>
  layout(
    title = "Top 10 Incident Frequency for Streetcar",
    xaxis = list(title = "Incident"),
    yaxis = list(title = "Frequency")
  )
```

### Subway
```{r echo=FALSE}
# Calculate frequency of each incident
top_10_incidents <- subway |>
  count(code_description, sort = TRUE) |>
  top_n(10, n)  # Select the top 10 incidents by frequency

# Plot the histogram for the top 10 incidents
plot_ly(x = reorder(top_10_incidents$code_description, -top_10_incidents$n),
        y = top_10_incidents$n, 
        type = "bar") |>
  layout(
    title = "Top 10 Incident Frequency for Subway",
    xaxis = list(title = "Incident"),
    yaxis = list(title = "Frequency")
  )
```



## Peak Times for Delays {.tabset}
Then, we investigate the peak times for delays in each of these transport types. As shown below, the frequency of delays for the subway and streetcars seems relatively consistent over time, except from around 2am to 5am, when it is usually closed and with slight peaks at 8am and 4pm, at peak hour. This led us to include a variable to indicate rush hour for our model. Moreover, the bus has a significant peak at 5pm, supporting this decision.

### Bus
```{r echo=FALSE}
bus$hms_time <- as_hms(bus$time)  # Convert to hms object first
bus$hour <- hour(bus$hms_time)  # Extract the hour from the time
bus$transport_type <- "Bus"

peak_time <- ggplot(bus, aes(x = hour)) +
  geom_bar(fill="orange") +
  labs(title = "Peak Times for Bus Delays", x = "Time", y = "Frequency of Delays") +
  theme_minimal()

ggplotly(peak_time)
```

### Streetcar
```{r echo=FALSE}
streetcar$hms_time <- as_hms(streetcar$time)  # Convert to hms object first
streetcar$hour <- hour(streetcar$hms_time)  # Extract the hour from the time
streetcar$transport_type <- "Streetcar"

peak_time <- ggplot(streetcar, aes(x = hour)) +
  geom_bar(fill="lightblue") +
  labs(title = "Peak Times for Streetcar Delays", x = "Time", y = "Frequency of Delays") +
  theme_minimal()

ggplotly(peak_time)
```
### Subway
```{r echo=FALSE}
subway$hms_time <- as_hms(subway$time)  # Convert to hms object first
subway$hour <- hour(subway$hms_time)  # Extract the hour from the time
subway$transport_type <- "Subway"

peak_time <- ggplot(subway, aes(x = hour)) +
  geom_bar(fill="lightgreen") +
  labs(title = "Peak Times for Subway Delays", x = "Time", y = "Frequency of Delays") +
  theme_minimal()

ggplotly(peak_time)
```

### All three

```{r echo=FALSE}
# Combine the datasets into one
combined_data <- bind_rows(subway, bus, streetcar)

# Summarize the counts by hour and transport type
hourly_counts_combined <- combined_data |>
  group_by(hour, transport_type) |>
  summarize(frequency = n(), .groups = 'drop')

# Plot the combined data with color to differentiate transport types
peak_time_combined <- ggplot(hourly_counts_combined, aes(x = hour, y = frequency, fill = transport_type)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(title = "Peak Times for Delays by Transport Type", x = "Time", y = "Frequency of Delays") +
  theme_minimal()

# Convert to interactive plot using plotly
ggplotly(peak_time_combined)
```


```{r echo=FALSE, eval=FALSE}
## Number of Delays by Date {.tabset}
# Count the number of delays for each date
delay_counts <- combined_data |>
  group_by(date) |>
  summarise(counts = n())

# Plot delays by date
delay_date <- ggplot(delay_counts, aes(x = date, y = counts)) +
  geom_point(stat = "identity", fill = "steelblue", color = "black", alpha = 0.7) +
  labs(
    title = "Number of Delays by Date",
    x = "Date",
    y = "Number of Delays"
  ) +
  theme_minimal()

ggplotly(delay_date)
```

## Delay Durations by Transport Type {.tabset}

After removing outliers from the data, we investigate how the distribution of the delay durations vary between the different transport types in the following figures. The violin plots suggest that the delay durations for buses has a large spread, as compared to streetcars and the subway. The subway seems to have a significant number of values near 0, while the streetcar has peaks at around 0 and 10 minutes. The animated figure shows how this varies over months: it seems that the number of delays increases in the summer months, although the general pattern remains consistent over the months. This motivated us to include season as a variable of interest in our models.

### By Transport Type

```{r echo=FALSE}
min_delay_counts_combined <- combined_data |>
  group_by(transport_type) |>
  mutate(Q1 = quantile(min_delay, 0.25),
         Q3 = quantile(min_delay, 0.75),
         IQR = Q3 - Q1) |>
  ungroup() |>
  filter(min_delay >= (Q1 - 1.5 * IQR) & min_delay <= (Q3 + 1.5 * IQR))

# Build the violin plot for `min_delay`
violin_plot <- ggplot(min_delay_counts_combined, aes(x = transport_type, y = min_delay, fill = transport_type)) +
  geom_violin(trim = FALSE) +
  labs(title = "Distribution of Delay Duration by Transport Type",
       subtitle = "TTC delays less than 40 minutes in 2024",
       x = "Transport Type", 
       y = "Delay (minutes)") +
  theme_minimal() +
  scale_fill_manual(values = c("Bus" = "lightblue", "Subway" = "lightgreen", "Streetcar" = "lightcoral"))

ggplotly(violin_plot)
```

### By Transport Type and Month
```{r echo=FALSE}
animated_plot <- combined_data %>% 
  filter(min_delay < 40) %>% 
  ggplot(aes(x=min_delay, fill = transport_type)) +
  geom_histogram() +
  facet_wrap(~transport_type) +
  transition_time(month(date)) +  # Animate over time
  ease_aes('linear') +
  labs(title = "Distribution of Delays by Transport Type in {month.name[frame_time]}",
     subtitle = "TTC delays less than 40 minutes in 2024",  # Add the date dynamically
     x = "Delay (minutes)", 
     y = "Count") +
  theme_bw() +
  theme(plot.title = element_text(face = "bold")) +
  theme(legend.position = "none")

animate(animated_plot, nframes = 100, renderer = gifski_renderer())
```